Heuristics and Really Hard Instances for Subgraph Isomorphism Problems

### 2016-05-11 poster/poster.tex
More postering

\medskip

@@

\block{The Expected Number of Solutions, and Heuristics}{
    If we randomly generate a pattern graph with $p$ vertices and density $d_p$, and a target graph
    with $t$ vertices and density $d_t$, the expected number of solutions to the non-induced
    isomorphism problem is \[
        \mathsf{\langle Sol \rangle = t \cdot (t - 1) \cdot \ldots \cdot (t - p + 1)\cdot {d_t}^{d_p \cdot
        \binom{p}{2}} } \textnormal{.} \]
}

@@

    We can also look at what happens if we vary both the pattern and the target density, for
    different orders of pattern. The top row shows satisfiability---the black lines show where the
    expected number of solutions is one. The bottom row shows the search cost for one of the three
    algorithms we consider in the paper. As expected, the ``really hard'' instances are near the
    phase transition.

@@

\block{Induced Phase Transitions}{
\begin{wrapfigure}[21]{r}{0.62\linewidth}\begin{center}
    \small\input{gen-graph-induced}
\end{center}\end{wrapfigure}

    For induced isomorphisms, the expected number of solutions is \[
        \mathsf{\langle Sol \rangle = t \cdot (t - 1) \cdot \ldots \cdot (t - p + 1)\cdot
                    {d_t}^{d_p \cdot \binom{p}{2}} \cdot {(1 - d_{t})}^{(1 - d_{p}) \cdot \binom{p}{2}} } \textnormal{.} \]

    \vspace*{52ex}
}

>>>>

\bigskip

@@

\block{The Expected Number of Solutions, and Heuristics}{

If we randomly generate a pattern graph with $p$ vertices and density $d_p$, and a target graph
with $t$ vertices and density $d_t$, the \textbf{expected number of solutions} to the non-induced
isomorphism problem is \[
    \mathsf{\langle Sol \rangle = t \cdot (t - 1) \cdot \ldots \cdot (t - p + 1)\cdot {d_t}^{d_p \cdot
    \binom{p}{2}} } \textnormal{.} \]

\bigskip

By considering when $\mathsf{\langle Sol \rangle = 1}$, we can \textbf{estimate the location} of the
satisfiable / unsatisfiable phase transition. (This is not entirely accurate due to variance,
particularly when the pattern is very sparse or very dense).

\bigskip

We can also use this formula to recover \textbf{variable} and \textbf{value ordering heuristics}: we
should make branching decisions which maximise the expected number of solutions during search. This
tells us to prefer small domains and pattern vertices of high degree, but target vertices of low
degree, which matches the empirically-determined heuristics used in state of the art solvers.

}

@@

    We can also look at what happens if we vary both the pattern and the target density, for
    different orders of pattern. The top row shows satisfiability---the black lines show where
    $\mathsf{\langle Sol \rangle = 1}$. The bottom row shows the search cost for one of the three
    algorithms we consider in the paper. As expected, the ``really hard'' instances are near the
    phase transition.

    \vspace*{0.9em}

@@

\block{Induced Phase Transitions}{
\begin{wrapfigure}[24]{r}{0.64\linewidth}\begin{center}
    \small\input{gen-graph-induced}
\end{center}\end{wrapfigure}

    For induced isomorphisms, the behaviour is much richer. We can derive an expected number of
    solutions of \[
        \mathsf{\langle Sol \rangle = t \cdot (t - 1) \cdot \ldots \cdot (t - p + 1)\cdot
                    {d_t}^{d_p \cdot \binom{p}{2}} \cdot {(1 - d_{t})}^{(1 - d_{p}) \cdot \binom{p}{2}} } \textnormal{.} \]

    \bigskip

    Again this predicts a sharp phase transition. When the pattern is small, we see hard instances
    near the phase transition, and easy instances elsewhere. However, when the pattern is larger,
    the central unsatisfiable region \textbf{remains very hard}, even though it is no longer near a
    phase transition.

    \bigskip

    This is not just a weakness of current subgraph isomorphism algorithms: the region is also hard
    for pseudo-boolean and boolean satisfiability solvers, and under reduction to the clique
    problem.

    \bigskip

    The central hard region \emph{is} predicted correctly by \textbf{constrainedness}, as we
    show in the third row of plots. Although far from a phase transition, this region is only
    slightly overconstrained.

    \bigskip

    Looking to maximise $\mathsf{\langle Sol \rangle}$ derives existing variable and value ordering
    heuristics, in the case that both the pattern and target graphs are sparse. The formula suggests
    that algorithms should swap heuristics when the pattern and target are dense---empirically, this
    does indeed give an improvement. However, maximising the expected number of solutions is
    \textbf{not enough to select the best heuristic} in two eighths of the search space (and
    constrainedness does not help).
}
